Problem Set 5

### 2019-11-26 content/assignment/05-problem-set.Rmd
Fix PS 5 tables

| Variable | Description |
| `$Price_h$ | price of house $h$ (in thousands of $) |
| $Bedrooms_h$ | number of bedrooms in house $h$ |
| $Baths_h$ | number of bathrooms in house $h$ |
| $Pool_h$ | $\begin{cases} =1 & \text{if house } h \text{ has a pool} \\ =0 & \text{if house } h \text{ does not have a pool} \end{cases}$ |
| $View_h$ |  $\begin{cases} =1 & \text{if house } h \text{ has a nice view} \\ =0 & \text{if house } h \text{ does not have a nice view} \end{cases}$ |

@@

| Variable | Description |
| `infrate` | infant mortality rate (deaths per 100 in population) |
| `lead` | $=1$ if city has lead water pipes, $=0$ if did not have lead pipes |
| `pH` | water pH |

@@

| Variable | Description |
| `Country` | Name of country |
| `ISO` | Code of country (good for plotting) |
| `econ_freedom` | Economic Freedom Index score (2016) from 1 (least) to 10 (most free) |
| `pol_freedom` | Political freedom index score (2018) from 1 (least) top 10 (most free) |
| `gdp_pc` | GDP per capita (2018 USD) |
| `continent` | Continent of country |

>>>>

| Variable | Description |
|----------|-------------|
| `$Price_h$ | price of house $h$ (in thousands of $) |
| $Bedrooms_h$ | number of bedrooms in house $h$ |
| $Baths_h$ | number of bathrooms in house $h$ |
| $Pool_h$ | $\begin{cases} =1 & \text{if house } h \text{ has a pool} \\ =0 & \text{if house } h \text{ does not have a pool} \end{cases}$ |
| $View_h$ |  $\begin{cases} =1 & \text{if house } h \text{ has a nice view} \\ =0 & \text{if house } h \text{ does not have a nice view} \end{cases}$ |

@@

| Variable | Description |
|----------}-------------}
| `infrate` | infant mortality rate (deaths per 100 in population) |
| `lead` | $=1$ if city has lead water pipes, $=0$ if did not have lead pipes |
| `pH` | water pH |

@@

| Variable | Description |
|----------|-------------|
| `Country` | Name of country |
| `ISO` | Code of country (good for plotting) |
| `econ_freedom` | Economic Freedom Index score (2016) from 1 (least) to 10 (most free) |
| `pol_freedom` | Political freedom index score (2018) from 1 (least) top 10 (most free) |
| `gdp_pc` | GDP per capita (2018 USD) |
| `continent` | Continent of country |

### 2019-11-26 static/homeworks/05-problem-set.Rmd
Fix PS 5 tables

---
title: "Problem Set 5"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Due by Thursday, December 5, 2019*

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/05-problem-set-pdf.pdf)
- [<i class="fab fa-markdown"></i> `Download Markdown`](/homeworks/05-problem-set.Rmd)

@@

## Question 1
In your own words, describe what the "dummy variable trap" means. What precisely is the problem, and what is the standard way to prevent it?

## Question 2
In your own words, describe what an interaction term is used for, and give an example. You can use any type of interaction to explain your answer.

## Question 3
In your own words, describe when and why using logged variables can be useful.

## Question 4
In your own words, describe when we would use an $F$-test, and give some example (null) hypotheses. Describe intuitively and specifically (no need for the formula) what exactly $F$ is trying to test for.

@@

### Part A
Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law, including the standard errors in parentheses below the coefficients. Define

@@

### Part B
How do we use the regression to find the average GDP Growth rate for common law countries? For civil law countries? For the difference?

### Part C
Looking at the coefficients, does there appear to be a statistically significant difference in average GDP Growth Rates between Civil and Common law countries?

### Part D
Is the estimate on the difference likely to be unbiased? Why or why not?

### Part E
Now using the same table above, reconstruct the regression equation if instead of Common Law, we had used:

@@

@@

### Part A
Suppose she runs the following regression:
$$\widehat{\text{Price}_h}=119.20+29.76 \, \text{Bedrooms}_h+24.09 \, \text{View}_h+14.06 \, (\text{Bedrooms}_h \times \text{View}_h)$$

> What does each coefficient mean?

### Part B
Write out **two** separate regression equations, one for houses *with* a nice view, and one for homes *without* a nice view. Explain each coefficient in each regression.

### Part C
Suppose she runs the following regression:

@@

> What does each coefficient mean?

### Part D
Find the expected price for:

@@

### Part E
Suppose she runs the following regression:
$$\widehat{\text{Price}_h}=87.90+53.94 \, \text{Bedrooms}_h+15.29 \, \text{Baths}_h+16.19 \, (\text{Bedrooms}_h \times \text{Baths}_h)$$

> What is the marginal effect of adding an additional *bedroom* if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?

### Part F
What is the marginal effect of adding an additional *bathroom* if the house has 1 bedroom? 2 bedrooms? 3 bedrooms?

@@

### Part A
Suppose we estimate the following simple model relating deviation in  temperature to year:

@@

> Interpret the coefficient on Year (i.e. $\hat{\beta_1})$

### Part B
Predict the (deviation in) temperature for the year 1900 and for the year 2000.

### Part C
Suppose we believe temperature deviations are increasing at an increasing rate, and introduce a quadratic term and estimate the following regression model:

@@

What is the marginal effect on (deviation in) global temperature of one additional year elapsing?

### Part D
Predict the marginal effect on temperature of one more year elapsing starting in 1900, and in 2000.

### Part E
Our quadratic function is a $U$-shape. According to the model, at what year was temperature (deviation) at its minimum?

@@

### Part A
Suppose we estimate the following simple regression:

@@

> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$

### Part B
Now suppose we estimate the regression using a linear-log model:

@@

> Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$

### Part C
Now suppose we estimate the regression using a log-linear model:

@@

> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$

### Part D
Now suppose we estimate the regression using a log-log model:

@@

> Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$

### Part E
Suppose we include several other variables into our regression and want to determine which variable(s) have the largest effects, a State's cell plans, population, or amount of miles driven. Suppose we decide to *standardize* the data to compare units, and we get:

@@

Interpret the coefficients on cell plans, population, and miles driven. Which has the largest effect on fatalities?

### Part F
Suppose we wanted to make the claim that it is *only* miles driven, and neither population nor cell phones determine traffic fatalities. Write (i) the null hypothesis for this claim and (ii) the estimated restricted regression equation.

### Part G
Suppose the $R^2$ on the original regression from (e) was 0.9221, and the $R^2$ from the restricted regression is 0.9062. With 50 observations, calculate the $F$-statistic.

@@

| Variable | Description |
|----------}-------------}
| `infrate` | infant mortality rate (deaths per 100 in population) |
| `lead` | $=1$ if city has lead water pipes, $=0$ if did not have lead pipes |
| `pH` | water pH |

@@

Using `R` to examine the data, find the average infant mortality rate for cities with lead pipes and for cities without lead pipes. Calculate the difference, and run a $t$-test to determine if this difference is statistically significant.

@@

Run a regression of `infrate` on `lead`, and write down the estimated regression equation. Use the regression coefficients to find:

@@

@@

Does the pH of the water matter? Include `ph` in your regression from part B. Write down the estimated regression equation, and interpret each coefficient (note there is no interaction effect here). What happens to the estimate on `lead`?

@@

The amount of lead leached from lead pipes normally depends on the chemistry of the water running through the pipes: the more acidic the water (lower pH), the more lead is leached. Create an interaction term between lead and pH, and run a regression of `infrate` on `lead`, `pH`, and your interaction term. Write down the estimated regression equation. Is this interaction significant?

@@

What we actually have are two different regression lines. Visualize this with a scatterplot between `infrate` $(Y)$ and `ph` $(X)$ by `lead`.

@@

Do the two regression lines have the same intercept? The same slope? Use the original regression in part D to test these possibilities.

@@

Take your regression equation from part D and rewrite it as two separate regression equations (one for no lead and one for lead). Interpret the coefficients for each.

@@

Double check your calculations in G are correct by running the regression in D twice, once for cities without lead pipes and once for cities with lead pipes.^[`filter()` the data first, then use the filtered data for the `data=` in each regression.]

@@

Use `huxtable` to make a nice output table of all of your regressions from parts B, C, and D.

@@

### Part A
Does economic freedom affect GDP per capita? Create a scatterplot of `gdp_pc` (`Y`) against `econ_freedom` (`x`). Does the effect appear to be linear or nonlinear?

### Part B
Run a simple regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. What is the marginal effect of `econ_freedom` on `gdp_pc`?

### Part C
Let's try a quadratic model. Run a quadratic regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation.

### Part D
Add the quadratic regression to your scatterplot.

### Part E
What is the marginal effect of `econ_freedom` on `gdp_pc`?

### Part F
As a quadratic model, this relationship should predict an` econ_freedom` score where `gdp_pc` is at a *minimum*. What is that minimum Economic Freedom score, and what is the associated GDP per capita?

### Part G
Run a cubic model to see if we should keep going up in polynomials. Write out the estimated regression equation. Should we add a cubic term?

### Part H
Another way we can *test* for non-linearity is to run an $F$-test on all non-linear variables - i.e. the quadratic term and the cubic term $(\hat{\beta_2}$ and $\hat{\beta_3}$) and test against the null hypothesis that:
$$H_0: \hat{\beta_2} = \hat{\beta_3} = 0$$

> Run this joint hypothesis test, and what can you conclude?

### Part I
Instead of a polynomial model, try out a logarithmic model. It is hard to interpret percent changes on an index, but it is easy to understand percent changes in GDP per capita, so run a **log-linear** regression. Write out the estimated regression equation. What is the marginal effect of `econ_freedom`?

### Part J
Make a scatterplot of your log-linear model with a regression line.

### Part K
Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.

>>>>

---
title: "Problem Set 5"
author: "YOUR NAME HERE" # put your name here! 
date: "ECON 480 - Fall 2019"
output: html_document # change to pdf if you'd like, this will make a webpage (you can email it and open in any browser)
---

<!--CLICK "KNIT" ABOVE TO RENDER TO HTML, PDF, OR WORD OUTPUT
In fact, try knitting right away and see what this produces!
-->

*Due by Thursday, November 21, 2019*

@@

## Question 1
**In your own words, describe what the "dummy variable trap" means. What precisely is the problem, and what is the standard way to prevent it?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

## Question 2
**In your own words, describe what an interaction term is used for, and give an example. You can use any type of interaction to explain your answer.**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

## Question 3

**In your own words, describe when and why using logged variables can be useful.**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

## Question 4

**In your own words, describe when we would use an $F$-test, and give some example (null) hypotheses. Describe intuitively and specifically (no need for the formula) what exactly $F$ is trying to test for.**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

@@

**Using the group means, write a regression equation for a regression of GDP Growth rate on Common Law, including the standard errors in parentheses below the coefficients. Define**

@@

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**How do we use the regression to find the average GDP Growth rate for common law countries? For civil law countries? For the difference?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Looking at the coefficients, does there appear to be a statistically significant difference in average GDP Growth Rates between Civil and Common law countries?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Is the estimate on the difference likely to be unbiased? Why or why not?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Now using the same table above, reconstruct the regression equation if instead of Common Law, we had used:**

@@

---

<!--WRITE YOUR ANSWERS BELOW -->

---

@@

**Suppose she runs the following regression:**
$$\widehat{\text{Price}_h}=119.20+29.76 \, \text{Bedrooms}_h+24.09 \, \text{View}_h+14.06 \, (\text{Bedrooms}_h \times \text{View}_h)$$

> **What does each coefficient mean?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Write out _two_ separate regression equations, one for houses _*_with_ a nice view, and one for homes _without_ a nice view. Explain each coefficient in each regression.**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Suppose she runs the following regression:**

@@

> **What does each coefficient mean?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

### Part D
**Find the expected price for:**

@@

---

<!--WRITE YOUR ANSWERS BELOW -->

---

### Part E
**Suppose she runs the following regression:**
$$\widehat{\text{Price}_h}=87.90+53.94 \, \text{Bedrooms}_h+15.29 \, \text{Baths}_h+16.19 \, (\text{Bedrooms}_h \times \text{Baths}_h)$$

> **What is the marginal effect of adding an additional _bedroom_ if the house has 1 bathroom? 2 bathrooms? 3 bathrooms?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

What is the marginal effect of adding an additional *bathroom* if the house has 1 bedroom? 2 bedrooms? 3 bedrooms?

---

<!--WRITE YOUR ANSWERS BELOW -->

---

@@

**Suppose we estimate the following simple model relating deviation in  temperature to year:**

@@

> **Interpret the coefficient on Year (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Predict the (deviation in) temperature for the year 1900 and for the year 2000.**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Suppose we believe temperature deviations are increasing at an increasing rate, and introduce a quadratic term and estimate the following regression model:**

@@

**What is the marginal effect on (deviation in) global temperature of one additional year elapsing?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Predict the marginal effect on temperature of one more year elapsing starting in 1900, and in 2000.**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Our quadratic function is a $U$-shape. According to the model, at what year was temperature (deviation) at its minimum?**
---

<!--WRITE YOUR ANSWERS BELOW -->

---

@@

**Suppose we estimate the following simple regression:**

@@

> **Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Now suppose we estimate the regression using a linear-log model:**

@@

> **Interpret the coefficient on ln(cell plans) (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Now suppose we estimate the regression using a log-linear model:**

@@

> **Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Now suppose we estimate the regression using a log-log model:**

@@

> **Interpret the coefficient on cell plans (i.e. $\hat{\beta_1})$**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Suppose we include several other variables into our regression and want to determine which variable(s) have the largest effects, a State's cell plans, population, or amount of miles driven. Suppose we decide to _standardize_ the data to compare units, and we get: **

@@

**Interpret the coefficients on cell plans, population, and miles driven. Which has the largest effect on fatalities?**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Suppose we wanted to make the claim that it is _only_ miles driven, and neither population nor cell phones determine traffic fatalities. Write (i) the null hypothesis for this claim and (ii) the estimated restricted regression equation.***

---

<!--WRITE YOUR ANSWERS BELOW -->

---

**Suppose the $R^2$ on the original regression from (e) was 0.9221, and the $R^2$ from the restricted regression is 0.9062. With 50 observations, calculate the $F$-statistic.**

---

<!--WRITE YOUR ANSWERS BELOW -->

---

@@

| Variable | Description |
|----------|-------------|
| `infrate` | infant mortality rate (deaths per 100 in population) |
| `lead` | $=1$ if city has lead water pipes, $=0$ if did not have lead pipes |
| `pH` | water pH |

@@

Part A

**Using `R` to examine the data, find the average infant mortality rate for cities with lead pipes and for cities without lead pipes. Calculate the difference, and run a $t$-test to determine if this difference is statistically significant.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**Run a regression of `infrate` on `lead`, and write down the estimated regression equation. Use the regression coefficients to find:**

@@

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**Does the pH of the water matter? Include `ph` in your regression from part B. Write down the estimated regression equation, and interpret each coefficient (note there is no interaction effect here). What happens to the estimate on `lead`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**The amount of lead leached from lead pipes normally depends on the chemistry of the water running through the pipes: the more acidic the water (lower pH), the more lead is leached. Create an interaction term between lead and pH, and run a regression of `infrate` on `lead`, `pH`, and your interaction term. Write down the estimated regression equation. Is this interaction significant?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**What we actually have are two different regression lines. Visualize this with a scatterplot between `infrate` $(Y)$ and `ph` $(X)$ by `lead`.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**Do the two regression lines have the same intercept? The same slope? Use the original regression in part D to test these possibilities.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**Take your regression equation from part D and rewrite it as two separate regression equations (one for no lead and one for lead). Interpret the coefficients for each.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**Double check your calculations in G are correct by running the regression in D twice, once for cities without lead pipes and once for cities with lead pipes.**^[`filter()` the data first, then use the filtered data for the `data=` in each regression.]

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**Use `huxtable` to make a nice output table of all of your regressions from parts B, C, and D.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

@@

**Does economic freedom affect GDP per capita? Create a scatterplot of `gdp_pc` (`Y`) against `econ_freedom` (`x`). Does the effect appear to be linear or nonlinear?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

**Run a simple regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. What is the marginal effect of `econ_freedom` on `gdp_pc`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

**Let's try a quadratic model. Run a quadratic regression of `gdp_pc` on `econ_freedom`. Write out the estimated regression equation. **

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

**Add the quadratic regression to your scatterplot.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

**What is the marginal effect of `econ_freedom` on `gdp_pc`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

**As a quadratic model, this relationship should predict an` econ_freedom` score where `gdp_pc` is at a *minimum*. What is that minimum Economic Freedom score, and what is the associated GDP per capita?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

**Run a cubic model to see if we should keep going up in polynomials. Write out the estimated regression equation. Should we add a cubic term?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

**Another way we can _test_ for non-linearity is to run an $F$-test on all non-linear variables - i.e. the quadratic term and the cubic term $(\hat{\beta_2}$ and $\hat{\beta_3}$) and test against the null hypothesis that:**
$$H_0: \hat{\beta_2} = \hat{\beta_3} = 0$$

> **Run this joint hypothesis test, and what can you conclude?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

**Instead of a polynomial model, try out a logarithmic model. It is hard to interpret percent changes on an index, but it is easy to understand percent changes in GDP per capita, so run a _log-linear_ regression. Write out the estimated regression equation. What is the marginal effect of `econ_freedom`?**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

### Part J

**Make a scatterplot of your log-linear model with a regression line.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```

---

### Part K

**Put all of your results together in a regression output table with `huxtable` from your answers in questions B, C, G, and H.**

---

<!--WRITE YOUR ANSWERS BELOW -->

```{r}
# write your code here
```
